---
title: "Bivariate Analysis"
author: "Mahendra Mariadassou, INRAE <br> .small[from original slides by Tristan Mary-Huard]"
date: "Shandong University, Weihai (CN)<br>Summer School 2024"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: ["css/custom_weihai_ss.css", default]
    lib_dir: libs
    nature:
      ratio: '16:9'
      slideNumberFormat: '%current%'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      before_body: macros.html
height: 1080
width: 1920
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,comment=NA,message=FALSE)
library('tidyverse')
library('gridExtra')
library(data.table)
library(scales)
```

## The AgroParisTech dataset

.question[quizz0]

```{r}
set.seed(42)
n <- 731
Agro <- tibble(Gender = sample(c("F", "M"), size = n, prob = c(0.7, 0.3), replace = TRUE), 
               Age    = sample(19:23, size = n, prob = 5:1, replace = TRUE), 
               Year   = case_when(
                 Age == 19 ~ "1A", 
                 Age >= 22 ~ "3A",
                 Age == 20 ~ sample(c("1A", "2A"), size = n, replace = TRUE), 
                 Age == 21 ~ sample(c("2A", "3A"), size = n, replace = TRUE), 
               ),
               Height = if_else(Gender == "M", rnorm(n = n, mean = 178, sd = 6), rnorm(n = n, mean = 165, sd = 4)), 
               Weight = runif(n, min = 18, max = 25) * (Height/100)^2 %>% round(2))
Agro <- Agro %>% select(all_of(c('Height','Weight','Age', 'Year', 'Gender')))
DT::datatable(Agro) %>% DT::formatRound(columns = c("Height", "Weight"), digits = 1)
```

---
class: middle, inverse, center

# Motivating example

---
## The AgroParisTech dataset

- 731 individuals

- 2 **qualitative** variables: Gender, Year

- 3 **quantitative** variables: Height, Weight, Age

--

How can we investigate the .blue[joint] distribution of 2 descriptors in a population ?  

--

.blue[3 kinds of joint analysis]:  

.pull-left[
- qualitative - qualitative   
- quantitative - qualitative  
- quantitative - quantitative  
]

.pull-right[
- Ex: Gender and Year  
- Ex: Height and Gender  
- Ex: Height and Weight  
]

---
class: middle, inverse, center

# Qualitative - Qualitative

---
## Qualitative - Qualitative

The couple (`Gender`,`Year`) is directly described through its joint distribution

$$P\left(G=g\bigcap Y=y \right) = \frac{n_{gy}}{n} \ \ .$$

The contingency table displays exhaustive information: .question[Quizz1]

.pull-left[
```{r}
Agro  |>  
  janitor::tabyl(Year, Gender) |> 
  janitor::adorn_totals(where = "col")  |> 
  janitor::adorn_totals(where = "row")
```
]

--

.col-right[
```{r}
Agro  |>  
  janitor::tabyl(Year, Gender) |> 
  janitor::adorn_totals(where = "col")  |> 
  janitor::adorn_totals(where = "row") |> 
  janitor::adorn_percentages(denominator = "all") |> 
  janitor::adorn_rounding(digits = 3)
```
]

--

.pull-left[
```{r}
Agro  |>  
  janitor::tabyl(Year, Gender) |> 
  janitor::adorn_totals(where = "col")  |> 
  janitor::adorn_percentages(denominator = "row") |> 
  janitor::adorn_rounding(digits = 3)
```
]

--

.col-right[
```{r}
Agro  |>  
  janitor::tabyl(Year, Gender) |> 
  janitor::adorn_totals(where = "row")  |> 
  janitor::adorn_percentages(denominator = "col") |> 
  janitor::adorn_rounding(digits = 3)
```
]

---
#### Graphical representations 

.pull-left[
- All counts
```{r, fig.height=3.5}
ggplot(Agro, aes(x = Gender, fill = Year)) + geom_bar() + scale_fill_viridis_d()
```
]

--

.col-right[
- Percentages proportional to area
```{r, fig.height=3.5, message=FALSE}
plotdata <- Agro |> count(Gender, Year) |> group_by(Gender) |> 
  mutate(width = sum(n), 
         ymax = cumsum(n)/width, 
         ymin = cumsum(c(0, head(n, -1)))/width) |> 
  ungroup()
x_data <- plotdata %>% distinct(Gender, width) %>% 
               mutate(xmax = cumsum(width),
                      xmin = cumsum(c(0, head(width, -1))))
plotdata <- plotdata %>% inner_join(x_data)
ggplot(plotdata, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = Year)) + 
  geom_rect(color = "gray20") + 
  scale_fill_viridis_d() + 
  scale_x_continuous(breaks = x_data %>% mutate(x = (xmax + xmin)/2) %>% pull(x), 
                     labels = x_data$Gender, name = "Gender")
# ggplot(Agro, aes(x = Gender, fill = Year)) + scale_fill_viridis_d()
```
]

--

.pull-left[
- Conditional probabilities (by Gender)
```{r, fig.height=3.5}
ggplot(Agro, aes(x = Gender, fill = Year)) + geom_bar(position = "fill") + scale_fill_viridis_d()
```
]

--

.col-right[
- Conditional probabilities (by Year)
```{r, fig.height=3.5}
ggplot(Agro, aes(x = Year, fill = Gender)) + geom_bar(position = "fill") + coord_flip()
```
]

---
class: middle, inverse, center

# Quantitative - Qualitative

---
## Quantitative - Qualitative

```{r}
TH <- theme(axis.text=element_text(size=15),
      axis.title=element_text(size=25),
      axis.title.y=element_blank(),
      plot.title=element_text(size=15),
      strip.text.x = element_text(size = 20),
      legend.text=element_text(size=20))

Hist1 <- ggplot(Agro, aes(x=Height))+
  geom_histogram(bins=30,color='black',fill='white')+
  facet_grid(. ~ Gender, scales = "free_x") +
  TH
# ggsave(paste0(Rep,'Slides/Figures/HistHeight1.pdf'))
Hist2 <- ggplot(Agro, aes(x=Height))+
  geom_histogram(bins=30,color='black',fill='white')+
  facet_grid(. ~ Gender) +
  TH
# ggsave(paste0(Rep,'Slides/Figures/HistHeight2.pdf'))
Hist3 <- ggplot(Agro, aes(x=Height, color=Gender, fill=Gender)) +
  geom_histogram(position="identity", alpha=0.5,bins=30) +
  TH
# ggsave(paste0(Rep,'Slides/Figures/HistHeight3.pdf'))
BP <- ggplot(Agro, aes(x=Gender,y=Height, color = Gender)) +
  geom_boxplot() + 
  geom_jitter(width = 0.2, )  +
  scale_color_discrete(guide = "none") + 
  TH
# ggsave(paste0(Rep,'Slides/Figures/BoxPlotHeight.pdf'))
```
Each level of variable `Gender` defines a sub-population, in which variable `Height` can be described.

.pull-left[
```{r, out.height="400px"}
set.seed(42)
BP
```
]
.pull-right[
```{r, out.height="400px"}
Hist1
```
]

- Same graphical tool as for 1 population, but...

---
## Quantitative - Qualitative

Each level of variable `Gender` defines a sub-population, in which variable `Height` can be described.

.pull-left[
```{r, out.height="400px"}
set.seed(42)
BP
```
]
.pull-right[
```{r, out.height="400px"}
Hist2
```
]

- Same graphical tool as for 1 population, but .blue[pay attention] to scaling effect...  

---
## Quantitative - Qualitative

Each level of variable `Gender` defines a sub-population, in which variable `Height` can be described.

.pull-left[
```{r, out.height="400px"}
set.seed(42)
BP
```
]
.pull-right[
```{r, out.height="400px"}
Hist3
```
]

- Same graphical tool as for 1 population, but .blue[pay attention] to scaling effect or .blue[avoid] them

---
class: middle, inverse, center

# Quantitative - Quantitative

---
## Quantitative - Quantitative

As for the univariate case, when dealing with continuous variables the joint distribution cannot be explored exhaustively.

--

**Nevertheless**, graphical representations can be produced:

```{r, fig.height=4, fig.width=6, fig.align='center'}
TH <- theme(axis.text=element_text(size=7),
      axis.title=element_text(size=12),
      plot.title=element_text(size=15),
      strip.text.x = element_text(size = 20),
      legend.text=element_text(size=20))

ggplot(Agro, aes(x=Height,y=Weight))+
  geom_point(size=0.8) +
  TH
```

--

The relationship between Height and Weight looks quite linear.

.blue[Question]: How can the linearity of the relationship be quantified ?

---
## Covariance

.blue[Definition:] The covariance $\sigma_{X,Y}$ between two quantitative variables $X$ and $Y$ is

$$\sigma_{X,Y} = \sum_{i}\sum_{j}\left(x_i-E(X)\right)\left(y_j-E(Y)\right)P\left(X=X_i,Y=y_j\right)$$


for quantitative discrete variables, and

$$\sigma_{X,Y} = \int_{x}\int_{y}\left(x-E(X)\right)\left(y-E(Y)\right)\times f_{X,Y}(x,y)dxdy$$

for continuous variables. .question[Quizz2]

--

.blue[Examples:]

- Covariance between Height and Weight: $\sigma_{H,W} = `r with(Agro, cov(Height, Weight))`$

- Covariance between Height (in cm)  and Weight: $\sigma_{H,W} = `r with(Agro, cov(100 * Height, Weight))`$

- Covariance between Weight and Age: $\sigma_{H,A} = `r with(Agro, cov(Weight, Age))`$

--

.blue[Conclusion:]  Scaling makes covariance difficult to interpret.

---
## Correlation

.blue[Definition:] The correlation $\rho_{X,Y}$ between two quantitative variables $X$ and $Y$ is .question[Quizz3]

$$\rho_{X,Y} = \frac{\sigma_{X,Y}}{\sigma_X \times \sigma_Y} \ \ .$$

--

Division by the standard deviation $\Rightarrow$ get rid of the scaling effect.  

--

.def[Property:]  $\rho_{X,Y} \in [-1,\ 1]$

- $\rho_{X,Y} \approx 1$ $\Rightarrow$ positive linear relationship between $X$ and $Y$,

- $\rho_{X,Y} \approx -1$ $\Rightarrow$ negative linear relationship between $X$ and $Y$,

- $\rho_{X,Y} \approx 0$ $\Rightarrow$ no linear relationship between $X$ and $Y$,

--

.blue[Examples:]

- Correlation between Height and Weight: $\rho_{H,W} = `r with(Agro, cor(Height, Weight))`$

- Correlaion between Height (in cm) and Weight: $\sigma_{H,W} = `r with(Agro, cor(100 * Height, Weight))`$

- Correlation between Height and Age: $\rho_{H,A} = `r with(Agro, cor(Height, Age))`$


---
## Intuition on the covariance (I)

.question[quizz4]

```{r, fig.align='center', fig.width=12, fig.height=6}
data(datasaurus_dozen, package = "datasauRus")
ggplot(datasaurus_dozen %>% filter(dataset %in% c("slant_up", "slant_down", "h_lines", "v_lines", "x_shape")), 
       aes(x = x, y = y, colour = dataset))+
  geom_point() +
  theme_void() +
  theme(legend.position = "none", strip.text = element_text(size = 20)) +
  facet_wrap(~dataset, ncol = 3)
```

---
## Intuition on the covariance (II)

.alert[All] dataset have the same summary statistics (and the same correlation $\rho = -0.06$) !!

```{r, fig.align='center', fig.width=12, fig.height=6}
data(datasaurus_dozen, package = "datasauRus")
ggplot(datasaurus_dozen %>% filter(dataset %in% c("slant_up", "slant_down", "h_lines", "v_lines", "x_shape", "dino")), 
       aes(x = x, y = y, colour = dataset))+
  geom_point() +
  theme_void() +
  theme(legend.position = "none", strip.text = element_text(size = 20)) +
  facet_wrap(~dataset, ncol = 3)
```

---
## About interpretation...

```{r,fig.height=5,fig.width=10,warning=FALSE, fig.align='center'}
set.seed(42)
PointSize=3
TH <- theme(axis.text=element_text(size=12),
      axis.title=element_text(size=15),
      plot.title=element_text(size=15),
      strip.text.x = element_text(size = 20),
      legend.text=element_blank())

TV <- c(13,20,23,25,27,31,36,46,55,63,70,76,81,85)
Nb <- c(8,8,9,10,11,11,12,16,18,19,20,21,22,23)
DF <- data.frame(TV,Nb)
GG1 <- ggplot(DF,aes(x=TV,y=Nb)) +
  geom_point(size=PointSize) +
  xlab('TV sold') +
  ylab('Nb autism cases') +
  annotate("text", x = 20, y=22, label = paste('Cor=',round(cor(TV,Nb),2)),
           color='red',size=5) +
  ggtitle('TV and mental disease') +
  TH

X_out <- rnorm(20,m=2,sd=2)
Y_out <- 5 -0.7*X_out + rnorm(20,m=0,sd=0.7)
qui <- which.max(X_out)
Y_out[qui] = 25
GG2 <- ggplot(data.frame(X_out,Y_out),aes(x=X_out,y=Y_out)) +
  geom_point(size=PointSize) +
  annotate("text", x = -0.2, y=24, label = paste('Cor=',round(cor(X_out,Y_out),2)),
           color='red',size=5) +
  labs(x = "X", y = "Y") + 
  xlim(c(min(-1,min(X_out)),max(6,max(X_out)))) +
  ylim(c(min(-0,min(Y_out)),max(25,max(Y_out)))) +
  ggtitle('Outlier') +
  TH

X <- rnorm(40,m=0,sd=1)
Y <- rnorm(40,m=0,sd=1)
Z <- c(rep(0,20),rep(1,20))
X[Z==1] = X[Z==1] + 5
Y[Z==1] = Y[Z==1] + 3
GG3 <- ggplot(data.frame(X,Y),aes(x=X,y=Y,color=Z)) +
  geom_point(size=PointSize, show.legend = FALSE) +
  annotate("text", x = -1, y=4, label = paste('Cor=',round(cor(X,Y),2)),
           color='red',size=5) +
  xlim(c(min(-2,min(X)),max(7,max(X)))) +
  ylim(c(min(-2,min(Y)),max(5,max(Y)))) +
  ggtitle("Batch effect") +
  TH

X <- runif(40,-3,3)
Y <- 5 -0.7*X^2 + rnorm(20,m=0,sd=0.4)
GG4 <- ggplot(data.frame(X,Y),aes(x=X,y=Y)) +
  geom_point(size=PointSize, show.legend = FALSE) +
  xlim(c(min(-3,min(X)),max(3,max(X)))) +
  ylim(c(min(-2,min(Y)),max(6,max(Y)))) +
  annotate("text", x = -2, y=5, label = paste('Cor=',round(cor(X,Y),2)),
           color='red',size=5) +
  ggtitle("Non linear relationship") +
  TH

gridExtra::grid.arrange(grobs=list(GG1,GG2,GG3,GG4) ,nrow = 2, as.table = FALSE)
```

.blue[Conclusion:]

- Correlation does not mean causality,

- Correlation .alert[does not replace] graphical representation.

---
## About the effect of the outlier 

```{r, fig.align='center', fig.width=12, fig.height=6}
GG2 <- GG2 + ggtitle("With outlier")

GG2_bis <- ggplot(tibble(X_out = X_out[-qui],Y_out = Y_out[-qui]),aes(x=X_out,y=Y_out)) +
  geom_point(size=PointSize) +
  annotate("text", x = -0.2, y=24, label = paste('Cor=',round(cor(X_out[-qui],Y_out[-qui]),2)),
           color='red',size=5) +
  labs(x = "X", y = "Y") + 
  xlim(c(min(-1,min(X_out[-qui])),max(6,max(X_out[-qui])))) +
  ylim(c(min(-0,min(Y_out[-qui])),max(25,max(Y_out[-qui])))) +
  ggtitle('Without outlier') +
  TH

gridExtra::grid.arrange(grobs=list(GG2, GG2_bis), nrow = 1, as.table = FALSE)
```



---
## Exercise

```{r}
Diploma <- c(5,-2,2,0,5)
Income <- c(36,14,21,16,30)
MeanD <- mean(Diploma)
MeanI <- mean(Income)
VarD <- mean((Diploma-MeanD)**2)
VarI <- mean((Income-MeanI)**2)
CovDI <- cov(Diploma,Income)*4/5
CorDI <- cor(Diploma,Income)*4/5
```


An economist investigates the relationship between education level and income in a small firm with 5 employees:

```{r}
tibble(Name = c("Engineer", "CAP", "DUT", "High School", "Msc"), Diploma, Income)
```

Compute the expectation and variance per variable, and the covariance and correlation between the education level and income.

--

.blue[Answer]

$\widehat{\mu}_I = 23.4, \ \widehat{\mu}_D = 2,\ \widehat{\sigma}^2_I = 70.24 \ (8.4), \ \widehat{\sigma}^2_D = 7.6\ (2.75)$

$\widehat{\sigma}_{I,D} = 22, \ \widehat{\rho}_{I,D} = 0.95$