---
title: "Tests"
author: "Mahendra Mariadassou, INRAE <br> .small[from original slides by Tristan Mary-Huard]"
date: "Shandong University, Weihai (CN)<br>Summer School 2023"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: ["css/custom_weihai_ss.css", default]
    lib_dir: libs
    nature:
      ratio: '16:9'
      slideNumberFormat: '%current%'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      before_body: macros.html
height: 1080
width: 1920
---

```{r setup, include=FALSE,}
knitr::opts_chunk$set(echo = FALSE,comment=NA,message=FALSE,warning = FALSE)
library(ggplot2)
library('gridExtra')
library(dplyr)
DisplayVect <- function(v,NbItemPerLine=6,Max=30,Round=3){
  lv <- length(v)
  Max <- min(lv, Max)
  v <- v[1:Max]
  for (ii in 1:(lv%/%NbItemPerLine)){
    cat(round(v[((ii-1)*NbItemPerLine+1):(ii*NbItemPerLine)],Round),'\n')
  }
  if (lv%%NbItemPerLine !=0){
    cat(round(v[(lv- lv%%NbItemPerLine + 1):lv],Round),'\n')
  }
}
```

```{r xaringan-scribble, echo=FALSE}
xaringanExtra::use_scribble()
```

---
class: middle, inverse, center

# Warm-up

## Introducing the Fisher distribution 

---

## Prerequesites: Fisher distribution

Let $Z_1$ and $Z_2$ be two positive random variables such that  

- $Z_1\sim\chi^2(n_1)$,  
- $Z_2\sim\chi^2(n_2)$,  
- $Z_1\perp Z_2$.  


Define
$$F = \frac{Z_1/n_1}{Z_2/n_2}$$
--

Then $F$ is said to follow a Fisher distribution with degrees of freedom $n_1$ and $n_2$.

One notes
$$F \sim \F(n_1,n_2)$$
---
  
## Example 1: Infection
  
```{r}
## Parameters
Alpha2 <- 0.05

## Read data
NbPhages <- c(2,2,2,2,3,3,2,4,1,1,5,1,3,1,1,5,1,1,4,2)
NbBactos <- length(NbPhages)
S = sum(NbPhages)
M = mean(NbPhages)
Lambda0 <- round(-log(0.1),4)

## Find the confidence interval using normal approx + plug-in
LambdaHat= M
quant.pois <- qpois(p = 1-Alpha2,lambda = NbBactos*Lambda0)
ccl <- 'rejects'
if (LambdaHat <= quant.pois/NbBactos){ ccl='does not reject' }
```

Back to the phage infection example. Assume now that if at least 90% of the bacteria colony is infected then the biologist applies a treatment to fight the infection. What decision rule should be applied ?

```{r}
DisplayVect(v = NbPhages,NbItemPerLine = 10,Round = 0)
```
  
.blue[Objective]
Perform a test to decide whether the proportion of infected bacteria is higher than $90\%$ or not.  
  
---
  
## Modeling
  
Denote $X_i$ the number of phages obtained for the $i^{th}$ bacterium.  

- One assumes that bacteria are independent:  
$X_1 \perp X_2 \perp \dots \perp X_{n},\quad\text{with } n= `r NbBactos`$ 
- Measurements are discrete
$X_1,..., X_{n}  \sim \mathcal{P}(\bullet), \text{ i.i.d}$
- The infection level is unknown
$X_1,..., X_{n} \sim \mathcal{P}(\lambda), \text{ i.i.d}$


---

## Hypotheses

Hypotheses *must* be formulated such that they *concern a parameter* of the model.  
Here the model is 

$$X_1,..., X_{n} \sim \mathcal{P}(\lambda), \text{ i.i.d}$$ 

Consequently the hypotheses should concern $\lambda$.

--

$$\begin{eqnarray*}
\hspace{-0.8cm}\text{Proportion of infected bacteria } \geq \text{90\%} &\Rightarrow& P(X>0)\geq 0.9 \\
&\Rightarrow& P(X=0)\leq 0.1 \\
&\Rightarrow& e^{-\lambda}\leq 0.1 \\
&\Rightarrow& \lambda\geq -\ln(0.1)  \\
&\Rightarrow& \lambda\geq \lambda_0 (=`r Lambda0`) \\
\end{eqnarray*}$$

-- 

Hence
$$H_0: \{\lambda\leq \lambda_0\} \quad \text{vs} \quad H_1: \{\lambda\geq \lambda_0\}$$

---
  
## Estimation
  
Starting point: derive the ML estimator for the quantity of interest.  

$$\begin{eqnarray*}
  Lik_\lambda(x_1,...,x_n) &=& \prod_{i=1}^n f_\lambda(x_i) \quad\text{(i.i.d. assumption)}\\
  &=& \prod_{i=1}^n \frac{\lambda^{x_i}}{x_i!}e^{-\lambda} \\
  \Rightarrow LLik_\lambda(x_1,...,x_n) &=& \log(\lambda)\sum_{i=1}^{n}x_i - \sum_{i=1}^{n}\log(x_i!) - n\lambda 
\end{eqnarray*}$$

.blue[Derivation]
$$\frac{\partial LLik_\lambda(x_1,...,x_n)}{\partial \lambda}=  \frac{1}{\lambda}\sum_{i=1}^n x_i -n$$
Setting the derivative at 0, one gets: $\widehat{\lambda} = \frac{1}{n}\sum_{i=1}^n x_i=\bar{x}$. 
  
  
---

## Decision rule

Should be of the form
$$\text{If } \hat{\lambda}\geq s \text{ then reject } H_0$$
One needs to choose threshold $s$.  


.blue[Type I error control]  
Find $s$ such that $P_{H_0}(\overline{X}\geq s)\leq \alpha$  
$\Rightarrow$ Requires the (possibly approximate) distribution of $\overline{X}$.  

One has:
$$\begin{eqnarray*}
n\overline{X} \sim \P(n\lambda) \quad \text{ and/or } \quad  \sqrt{n}\frac{\overline{X}-\lambda}{\lambda} \overset{approx}{\sim} \N(0,1)
\end{eqnarray*}$$

.blue[Under $H_0$]:  
One has $\lambda \in [0,\lambda_0]$, still the $H_0$ distribution of $\overline{X}$ is not fully known...


---

## Worst case scenario

Which value of $\lambda \in [0,\lambda_0]$ leads to the worst (i.e. the max) value of $P_{H_0}(\overline{X}\geq s)$ ?  


```{r,fig.height=6}
## Parameters
MinLambda = 1
MaxLambda = - log(0.1)
NbVal = 3
MaxX <- 15

## Theme
Coef=1.2  
THFtn <- function(Coef){
  theme(axis.text=element_text(size=10*Coef),
      axis.title=element_text(size=15*Coef),
      plot.title=element_text(size=15*Coef),
      strip.text.x = element_text(size = 20*Coef),
      legend.text=element_text(size=12*Coef),
      legend.title = element_text(size=15*Coef))
} 
TH <- THFtn(Coef)
  
## Plot the risk
LambdaSet <- round(seq(MinLambda,MaxLambda,length.out = NbVal),4) %>% rep(.,each=MaxX+1)
Runs <- rep(0:MaxX,times=3)
Probs <- dpois(Runs,lambda = LambdaSet)
DF <- data.frame(LambdaSet,Runs,Probs,Lambda=as.character(LambdaSet)) 
ggplot(DF,aes(x=Runs,y=Probs,col=Lambda)) +
  geom_line(size=2) + 
  ylab('Prob(X=x)') + xlab('x') + ggtitle('Poisson distribution with different lambdas') +
  geom_vline(xintercept = 4.5,size=2) + TH
```

--

Worst value: $\lambda =\lambda_0$ !

---

## Back to the decision rule

```{r, echo = FALSE, eval = FALSE}
GG <- ggplot(data.frame(x = c(0, 15)), aes(x)) +
  stat_function(fun = dpois,args = list(lambda=Lambda0),size=2) +
  stat_function(fun = dpois,args = list(lambda=Lambda0),
                xlim = c(quant.pois,15),
                geom = "area",fill=2) +
  ylab('Value') + xlab('Density') + ggtitle(paste('Poisson(lambda0) distribution')) +
  TH
```


Considering the worst case scenario,
one looks for $s$ such that
$$P_{\lambda_0}\left( \overline{X}\geq s \right)\leq \alpha \Rightarrow P_{\lambda_0} \left( n\overline{X}\geq ns \right)\leq \alpha$$
Choose $ns = q_{n\lambda_0,1-\alpha} \Rightarrow s = \frac{q_{n\lambda_0,1-\alpha}}{n}$,

.blue[Application]
- $\widehat{\lambda}=$   `r LambdaHat`
- $n=$ `r NbBactos`
- $\alpha= 0.05$
- $q_{n\lambda_0,1-\alpha}=$ `r quant.pois` $\Rightarrow s=$ `r quant.pois/NbBactos`
--

.blue[Conclusion ?]

One `r ccl` the hypothesis that the proportion of infected bacteria is lower than 90\%


---

## Power of a test procedure

Recall that the test procedure is design such that:

- T1E is controlled at a given level,
- T2E is minimum given T1E is controlled.


T2E is minimum $\Rightarrow$ $P_{H_1}( \text{accept } H_0)$ is minimum $\Rightarrow$ $P_{H_1}(\text{reject} H_0)$ is maximum. 


$P_{H_1}(\text{reject } H_0)$ is the ability to reject $H_0$ when it false.
This key quantity is called the __power__ of the test procedure.

